---
title: "BIS581v"
output:
  word_document: default
  html_notebook: default
---

```{r}
library(tidyr)
library(tidyverse)
#load from CSV
centralData <- read.csv("Orders_Central.csv", header=TRUE)
westData <- read.csv("orders_west.csv", header=TRUE)
eastData <- read.delim("Orders_East.txt", header=TRUE, sep='\t')


# lowercase column names 
centralData <- centralData %>% rename_with(tolower)
westData <- westData %>% rename_with(tolower)
eastData <- eastData %>% rename_with(tolower)

# check column names
colnames(centralData)
colnames(westData)
colnames(eastData)

# Add the Region column for 'Central Data'
centralData <- centralData %>% mutate(region = "central")

# rename 'product' to 'product name'
centralData <- centralData %>% rename(product.name = product)

# Combine year, month, and day into proper Date columns
centralData <- centralData %>%
  mutate(order.date = as.Date(paste(order.year, order.month, order.day, sep = "-"), format = "%Y-%m-%d"),
         ship.date = as.Date(paste(ship.year, ship.month, ship.day, sep = "-"), format = "%Y-%m-%d"))

# change data type
centralData <- centralData %>%
  mutate(discounts = as.numeric(discounts))

# Remove columns
centralData <- centralData %>% select(-order.year, -order.month, -order.day, -ship.year, -ship.month, -ship.day)



# remove unneccessary column for west data
westData <- westData %>% select(-starts_with("right_"))

# Rename columns
westData <- westData %>%
  rename(discounts = discount)

# Rename columns 
eastData <- eastData %>%
  rename(discounts = discount)


# Standardize `order.date` and `ship.date` as Date objects in all datasets
centralData <- centralData %>%
  mutate(order.date = as.Date(order.date, format = "%Y-%m-%d"),
         ship.date = as.Date(ship.date, format = "%Y-%m-%d"))

# Correctly parse order.date and ship.date in westData
westData <- westData %>%
  mutate(order.date = as.Date(order.date, format = "%Y-%m-%d %H:%M:%S"),
         ship.date = as.Date(ship.date, format = "%Y-%m-%d %H:%M:%S"))

eastData <- eastData %>%
  mutate(order.date = as.Date(order.date, format = "%m/%d/%Y"),
         ship.date = as.Date(ship.date, format = "%m/%d/%Y"))

#convert Sales column in eastData
eastData <- eastData %>%
  mutate(sales = as.numeric(gsub("USD","",sales)))

# Merge the datasets
merged_data <- bind_rows(centralData, westData, eastData)
```

For this assignment, answer the following questions, but please note: the data may or may not be setup such that you can just answer. You may need to perform some "data wrangling" before you can answer. Include the code you use for any wrangling/EDA in your submission. Also include the code you use to answer each question along with text indicating what the answer is. You must provide proof of your answer by showing your R code. You may use online resources and talk to others, but what you submit must be your own work.

# 1.which region, on average, ships products faster:

-   The **West region** has the fastest average shipping time of **3.90 days** and similarly **East** with **3.91 days** .
-   The **Central region** has the slowest average shipping time of **4.06 days**.

```{r}
merged_data <- merged_data %>%
  mutate(time.to.ship = as.numeric(difftime(ship.date,order.date,units = "days")))

# Remove rows with missing or blank regions
merged_data <- merged_data %>%
  filter(region != "" & !is.na(region))

average_shipping_time <- merged_data %>%
  group_by(region) %>%
  summarise(avg_shipping_time = mean(time.to.ship, na.rm = TRUE)) %>%
  arrange(avg_shipping_time)

print(average_shipping_time)
```

# 2.Which products ship slowest by region:

-   In the **East region**, the product with the slowest shipping time is **Acme Elite Stainless Steel Scissors**, with an average shipping time of **7 days**.
-   In the **West region**, the product with the slowest shipping time is **3M Office Air Cleaner**, with an average shipping time of **7 days**.
-   In the **Central region**, the product with the slowest shipping time is **#6 3/4 Gummed Flap White Envelopes**, with an average shipping time of **7 days**.

```{r}
slowest_shipping_products <- merged_data %>%
  group_by(region, product.name) %>%
  summarise(avg_shipping_time = mean(time.to.ship, na.rm = TRUE), .groups = "drop") %>%
  group_by(region) %>%
  slice_max(avg_shipping_time, n = 1, with_ties = FALSE)

print(slowest_shipping_products)
```

# 3.Plot time to ship by category, by year.

-   The **Technology category** consistently had the longest average shipping time over the years, peaking at **4.2 days** in 2017 before dropping significantly in 2018.
-   The **Furniture category** showed the shortest average shipping times across most years.
-   The **Office Supplies category** maintained moderate shipping times, with minimal variation compared to other categories.

```{r}
#3.Plot time to ship by category, by year.

merged_data = merged_data %>% 
  mutate(order_year = as.numeric(format(order.date,"%Y")))

time_to_ship_by_category_year <- merged_data %>%
  group_by(category,order_year) %>%
  summarise(avg_shipping_time = mean(time.to.ship, na.rm = TRUE), .groups = "drop")

ggplot(time_to_ship_by_category_year, aes(x = order_year, y = avg_shipping_time, color = category)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Time to Ship by Category, by Year",
       x = "Year",
       y = "Average Time to Ship (days)",
       color = "Category") +
  theme_minimal()
```

# 4.which categories have highest profit by region, chain-wide?

-   The **Office Supplies category** generated the highest total profit of **217,839.04**, making it the most profitable category.
-   The **Technology category** follows with a total profit of **197,213.14**.
-   The **Furniture category** has the lowest total profit at **23,908.98**.

```{r}
# Calculate total profit by region and category
highest_profit_categories <- merged_data %>%
  group_by(region, category) %>%
  summarise(total_profit = sum(profit, na.rm = TRUE), .groups = "drop") %>%
  arrange(region, desc(total_profit)) %>%
  group_by(region) %>%
  slice_max(total_profit, n = 1, with_ties = FALSE)

# Calculate total profit by category (chain-wide)
highest_profit_chain_wide <- merged_data %>%
  group_by(category) %>%
  summarise(total_profit = sum(profit, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_profit)) %>%
  group_by(category) %>%
  slice_max(total_profit, n = 1, with_ties = FALSE)

print(highest_profit_categories)

print(highest_profit_chain_wide)
```

# 5.which segments have the lowest profit by region?

-   In the **East region**, the segment with the lowest profit is **Corporate**, with a total profit of **22,716.63**.
-   In the **West region**, the segment with the lowest profit is **Home Office**, with a total profit of **33,761.17**.
-   In the **Central region**, the segment with the lowest profit is **Consumer**, with a total profit of **8,564.05**.

```{r}
# Calculate total profit by region and segment
lowest_profit_segments <- merged_data %>%
  group_by(region, segment) %>%
  summarise(total_profit = sum(profit, na.rm = TRUE), .groups = "drop") %>%
  arrange(region, total_profit) %>%
  group_by(region) %>%
  slice_min(total_profit, n = 1, with_ties = FALSE)

# View the results
print(lowest_profit_segments)
```

# 6.What are yearly sales by region?

-   Sales in the **East region** grew steadily from 128,092.7 at the year-end of 2015 to 211,779.5 in 2018

-   Growth was rapid in the **West region** (from 442,618.9 sales 2015 to 773,738.5 sales 2018).

-   The **Central region** had stable sales, with 103,838.2 in 2015 and 102,874.2 in 2016.

```{r}
yearly_sales_by_region <- merged_data %>%
  group_by(region, order_year) %>%
  summarise(total_sales = sum(sales, na.rm = TRUE), .groups = "drop") %>%
  arrange(region, order_year)

# View the results
print(yearly_sales_by_region)
```
